---
title: "Indianapolis Metro Police Department Use of Force Data"
author: "Will Schrepferman"
date: "6/1/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = FALSE)
library(tidyverse)
library(dplyr)
library(janitor)
library(gt)
library(lubridate)
library(cowplot)
library(plotly)
library(scales)
```

## Introduction

This project uses public data to analyze the Indiana Metro Police Department's use of force since 2014. I used data from the OpenIndy Data Portal, part of Mayor Joe Hogsett's 'Disclose Indy' initative to ensure public access to Indianapolis' public records. No law requires the disclosure of police data across the state of Indiana as a whole.

## A Few Notes on Data Cleaning

Specifically, I used the following source: http://data.indy.gov/datasets/impd-complaints/data.

A few notes on structural changes made to filter data: data was downloaded directly from the site, but needed to be reformatted. Duplicate rows were removed, as it appears multiple reports were sometimes filed about the same instance (OpenIndy and the IMPD were contacted to get clarification on this, but the data will remain as is until information is received proving these assumptions wrong). Also, it appears that every type of force is counted as an instance- for example, if the same officer uses both a taser and physical pressure on the same citizen, that is logged as two separate instances. As mentioned before, this structural organization of the data was maintained, as it seemed critical to include every instance that officers saw fit to log.

I also re-named several columns in order to more directly convey their meaning. All told, there are 20,623 relevant instances in the data, which range from the beginning of 2014 to May 27, 2020 (I downloaded the data on June 1). Each instance has information on: the date, IMPD division, district, disposition, street number and name, type of force used, reason for use of force, circumstances of the officer's involvement (service), details on the arrest, injury, or hospitalization of the citizen, injury or hospitalization of the officer,  demographic information on both parties, and the officer's ID number.

Even with this cleaning, many issues remain with the data itself- thousands of instances are incompletely reported, analysis of the data over time indicates that data hasn't been fully reported from year to year (as can be seen in the 'Use of Force Over Time' section). Although this data is incomplete and flawed, though, it is still useful to conduct aggregate analysis in order to identify systemic trends.

```{r read_data}
raw_data <- read.csv("IMPD_Use_Of_Force.csv") %>%
  clean_names()
  
# rename columns and change data types
  
clean_data <- raw_data %>%
  mutate(incident_num = incnum) %>%
  mutate(date = as.Date(occurred_dt)) %>%
  mutate(division = as.character(udtext24a)) %>%
  mutate(district = as.character(udtext24b)) %>%
  mutate(disposition = as.character(disposition)) %>%
  mutate(street_num = as.numeric(street_n)) %>%
  mutate(street_name = as.character(street)) %>%
  mutate(type_of_force = as.character(uof_force_type)) %>%
  mutate(reason_for_force = as.character(uof_reason)) %>%
  mutate(service = as.character(service_type)) %>%
  mutate(citizen_arrested = as.character(cit_arrested)) %>%
  mutate(citizen_weapon = as.character(cit_weapon_type)) %>%
  mutate(citizen_injured = as.character(cit_injured)) %>%
  mutate(citizen_hospitalized = as.character(cit_hospital)) %>%
  mutate(officer_injured = as.character(off_injured)) %>%
  mutate(officer_hospitalized = as.character(off_hospital)) %>%
  mutate(citizen_race = as.character(race)) %>%
  mutate(citizen_sex = as.character(sex)) %>%
  mutate(officer_number = as.character(offnum)) %>%
  mutate(officer_race = as.character(off_race)) %>%
  mutate(officer_sex = as.character(off_sex)) %>%
  mutate(officer_age = as.character(off_age)) %>%
  mutate(officer_years_employed = as.numeric(off_yr_employ)) %>%
  select(incident_num, date, officer_number, division, district, disposition, street_num, street_name, type_of_force, reason_for_force, service, citizen_arrested, citizen_weapon, citizen_injured, citizen_hospitalized, officer_injured, officer_hospitalized, citizen_race, citizen_sex, officer_race, officer_sex, officer_age, officer_years_employed)

# remove duplicate rows

clean_data <- unique(clean_data) 

```

Here is a sample of rows from the data:

```{r display_data, include = TRUE}
sample_data <- clean_data %>%
  head(5) %>%
  gt()
```

## Overview

https://citybase-cms-prod.s3.amazonaws.com/d95654f9a1234d87afd0724cc65172c9.pdf
```{r impd_demo_bar}
#DEMOGRAPHICS OF IMPD

impd_demo_data <- tibble(race = c("Asian", "Hispanic", "Black", "White"), percent = c(0.1, 3.0, 16.0, 79.6))
impd_demo_bar <- impd_demo_data %>%
  ggplot(aes(x = reorder(race, percent), y = percent, fill = race)) +
  geom_bar(stat = "identity") +
  theme(legend.position = "none") +
  geom_text(aes(label=percent), hjust=-0.15, size=3.5) + 
  labs(x = "", title = "Racial Demographics of IMPD Force", y = "Percent of IMPD Force") +
  coord_flip()
```

```{r indy_demo_bar, include = TRUE}
# DEMOGRAPHICS OF INDIANAPOLIS

indy_demo_data <- tibble(race = c("Asian", "Black", "Hispanic", "White"), percent = c(3.2, 28.3, 10.2, 61.4))
indy_demo_bar <- indy_demo_data %>%
  ggplot(aes(x = reorder(race, percent), y = percent, fill = race)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label=percent), hjust=-0.15, size=3.5) + 
  ylim(0, 82) +
  theme(legend.position = "none") +
  labs(x = "", y = "Percent of Indianapolis Population", title = "Racial Breakdown of Indianapolis") +
  coord_flip()
```

https://www.census.gov/quickfacts/fact/table/indianapoliscitybalanceindiana/PST045219#qf-headnote-a



```{r use_of_force_demo, include = TRUE}
total_obs <- 20623

# DEMOGRAPHICS OF USE OF FORCE

race_overview_uof_plot <- clean_data %>%
  group_by(citizen_race) %>%
  summarize(count = n()) %>%
  filter(citizen_race == "Asian" | citizen_race == "Black" | citizen_race == "Hispanic" | citizen_race == "White") %>%
  mutate(percent = round(100*(count/total_obs), 1)) %>%
  mutate(citizen_race = factor(citizen_race, levels = c("Asian", "Hispanic", "Black", "White"))) %>%
  ggplot(aes(x = citizen_race, y = percent, fill = citizen_race)) +
  geom_bar(stat = "identity") +
  theme(legend.position = "none") +
  geom_text(aes(label=percent), hjust=-0.15, size=3.5) + 
  ylim(0, 70) +
  labs(x = "", title = "IMPD Uses of Force By Race", y = "Percent") +
  coord_flip()

# Indianapolis dempgraphics modified to fit scale of previous plot

indy_demo_bar2 <- indy_demo_data %>%
  ggplot(aes(x = reorder(race, percent), y = percent, fill = race)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label=percent), hjust=-0.15, size=3.5) + 
  ylim(0, 70) +
  theme(legend.position = "none") +
  labs(x = "", y = "Percent of Indianapolis Population", title = "Racial Breakdown of Indianapolis") +
  coord_flip()

```

```{r time_bar, include = TRUE}
yearly <- clean_data %>%
  filter(citizen_race == "Black" | citizen_race == "White" | citizen_race == "Asian" | citizen_race == "Hispanic") %>%
  mutate(year = year(date)) %>%
  filter(year <= 2019) %>%
  group_by(year, citizen_race) %>%
  summarize(count = n())


ggplot(yearly, aes(x = year, y = count, fill = citizen_race)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Instances of Force over Time by Race", x = "Year", y = "No. of instances")
```



```{r types_of_force, include = TRUE}
use_of_force_chart <- clean_data %>%
  filter(type_of_force != "") %>%
  group_by(type_of_force) %>%
  summarize(count = n(), na.rm = TRUE) %>%
  ggplot(aes(x = reorder(type_of_force, -count), y = count)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Type of Force Used", y = "Count", title = "IMPD Uses of Force by Type") +
  geom_text(aes(label=count), vjust=.33, hjust = -0.3, size=2.5, angle = 90) +
  ylim(0, 4700)

#508 instances where type not reported
```

```{r use_of_force}
uof_data <- clean_data %>%
  filter(type_of_force != "") %>%
  group_by(type_of_force) %>%
  summarize(count = n(), na.rm = TRUE) %>%
  arrange(-count)

# takes number i in list of types of force and outputs graph for that type

uof_race_function <- function(i) {
  type <- uof_data[i, 1] %>%
    slice(1) %>%
    pull(1)
  count <- uof_data[i, 2] %>%
    slice(1) %>%
    pull(1)
  
  specific_data <- clean_data %>%
    filter(type_of_force == type) %>%
    group_by(citizen_race) %>%
    summarize(count = n()) %>%
    filter(citizen_race == "Black" | citizen_race == "White" | citizen_race == "Asian" | citizen_race == "Hispanic")

  asian_add <- c("Asian", 0)
  black_add <- c("Black", 0)
  hispanic_add <- c("Hispanic", 0)
  white_add <- c("White", 0)
  
  new_data <- specific_data
  
  if ("Asian" %in% new_data$citizen_race == FALSE){
    new_data <- rbind(new_data, asian_add)
  }
  if ("Black" %in% new_data$citizen_race == FALSE){
    new_data <- rbind(new_data, black_add)
  }
  if ("Hispanic" %in% new_data$citizen_race == FALSE){
    new_data <- rbind(new_data, hispanic_add)
  }
  if ("White" %in% new_data$citizen_race == FALSE){
    new_data <- rbind(new_data, white_add)
  }
  
   new_data <- new_data %>%
    mutate(citizen_race = factor(citizen_race, levels = c("Asian", "Black", "Hispanic", "White")))
  
  a <- new_data[1,2] %>%
    pull(1)
  b <- new_data[2,2] %>%
    pull(1)
  h <- new_data[3,2] %>%
    pull(1)
  w <- new_data[4,2] %>%
    pull(1)
  
  a <- as.numeric(a)
  b <- as.numeric(b)
  h <- as.numeric(h)
  w <- as.numeric(w)
  
  total <- a + b + h + w
  
  graph <- new_data %>%
    ggplot(aes(x = citizen_race, y = as.numeric(count), fill = citizen_race)) +
    geom_bar(stat = "identity") +
    geom_segment(aes(x = 0.5, xend = 1.5, y = total*.003, yend = total*.003)) +
    geom_segment(aes(x = 1.5, xend = 2.5, y = total*.556, yend = total*.556)) +
    geom_segment(aes(x = 2.5, xend = 3.5, y = total*.046, yend = total*.046)) +
    geom_segment(aes(x = 3.5, xend = 4.5, y = total*.395, yend = total*.395)) +
    labs(title = paste("Force Type: ", type, "\nRacial Breakdown (", count, " instances)", sep = ""), x = "Citizen Race", y = "Count")
  
  print(graph)
}


g1 <- uof_race_function(1)
g2 <- uof_race_function(2)
g3 <- uof_race_function(3)
g4 <- uof_race_function(4)
g5 <- uof_race_function(5)
g6 <- uof_race_function(6)
g7 <- uof_race_function(7)
g8 <- uof_race_function(8)
g9 <- uof_race_function(9)
g10 <- uof_race_function(10)
g11 <- uof_race_function(11)
g12 <- uof_race_function(12)
g13 <- uof_race_function(13)
g14 <- uof_race_function(14)
g15 <- uof_race_function(15)
g16 <- uof_race_function(16)
g17 <- uof_race_function(17)
g18 <- uof_race_function(18)

types_grid <- plot_grid(g1, g2, g3, g4, g5, g6, g7, g8, g9, g10, g11, g12, g13, g14, g15, g16, g17, g18, nrow = 6, ncol = 6)

```


```{r data_analysis, include = TRUE}

# CALENDAR HEATMAP CREATOR

# get data from every day

daily <- clean_data %>%
  mutate(Date = date) %>%
  select(Date) %>%
  group_by(Date) %>%
  summarize(Instances = n())

daily_data <- daily %>%
  mutate(year = year(Date),
         month = month(Date, label = TRUE),
         wkday = fct_relevel(wday(Date, label=TRUE),
                             c("Mon", "Tue","Wed","Thu","Fri","Sat","Sun")),
         day = day(Date),
         wk = format(Date, "%W")) %>%
  filter(!is.na(year)) %>%
  select(Date, year, month, wkday, day, wk, Instances) %>%
  ggplot(aes(wk, wkday, fill = Instances, Date = Date)) +
  geom_tile() +
  scale_fill_gradient(low = "lightgreen", high = "orange") +
  geom_text(aes(label=Instances), size = 3) +
  labs(x = "", y = "", title = "Calendar Map of Use of Force", fill = "Instances", subtitle = "Number on Square is How Many Instances Were Used That Day") +
  theme(panel.background = element_blank(),
            axis.ticks = element_blank(),
            axis.text.x = element_blank(),
            strip.background = element_rect("grey92")
            ) +
     facet_grid(year~month, scales="free", space="free")

daily_data_interactive <- ggplotly(daily_data, tooltip = c("Date", "fill"))

```

```{r racial_analysis, include = TRUE}

# OFFICER/CITIZEN RACE HEATMAPS

# heatmap of observed race matrix

racial_data <- clean_data %>%
  select(citizen_race, officer_race) %>%
  filter(citizen_race == "Black" | citizen_race == "White" | citizen_race == "Asian" | citizen_race == "Hispanic") %>%
  filter(officer_race == "Black" | officer_race == "White" | officer_race == "Asian" | officer_race == "Hispanic") %>%
  group_by(citizen_race, officer_race) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  add_row(citizen_race = "Asian", officer_race = "Asian", count = 0) %>%
  add_row(citizen_race = "Hispanic", officer_race = "Asian", count = 0)

observed_plot <- racial_data %>%
  ggplot(aes(x = citizen_race, y = officer_race, fill = count, color = "black")) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "orange2") +
  geom_text(aes(label=count), size = 4, color = "black") + 
  theme(legend.position = "none") + 
  labs(x = "Citizen Race", y = "Officer Race")


racial_data <- racial_data %>%
  arrange(citizen_race, officer_race)

b <- 20264

# Baseline matrix if distribution is assumed to be even across general use of force data

racial_baseline_uof <- tibble(citizen_race = c("Asian", "Asian", "Asian", "Asian",
                                               "Black", "Black", "Black", "Black",
                                               "Hispanic", "Hispanic", "Hispanic", "Hispanic",
                                               "White", "White", "White", "White"),
                              officer_race = c("Asian", "Black", "Hispanic", "White",
                                               "Asian", "Black", "Hispanic", "White",
                                               "Asian", "Black", "Hispanic", "White",
                                               "Asian", "Black", "Hispanic", "White"),
                              count = c(.000003*b, .00048*b, .00009*b, .0024*b,
                                        .00056*b, .089*b, .00736*b, .443*b,
                                        .000046*b, .00736*b, .01668*b, .0366*b,
                                        .000395*b, .0632*b, .0119*b, .3144*b)
                              ) %>%
    mutate(count = round(count, 1))


                            

uof_base_plot <- racial_baseline_uof %>%
  ggplot(aes(x = citizen_race, y = officer_race, fill = count, color = "black")) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "orange2") +
  geom_text(aes(label=count), size = 4, color = "black") + 
  theme(legend.position = "none") + 
  labs(x = "Citizen Race", y = "Officer Race", title = "Use of Force Distribution Baseline")


# Baseline matrix if distribution is assumed to be even across general Indianapolis population


racial_baseline_indy <- tibble(citizen_race = c("Asian", "Asian", "Asian", "Asian",
                                               "Black", "Black", "Black", "Black",
                                               "Hispanic", "Hispanic", "Hispanic", "Hispanic",
                                               "White", "White", "White", "White"),
                              officer_race = c("Asian", "Black", "Hispanic", "White",
                                               "Asian", "Black", "Hispanic", "White",
                                               "Asian", "Black", "Hispanic", "White",
                                               "Asian", "Black", "Hispanic", "White"),
                              count = c(.000032*b, .00512*b, .00096*b, .0255*b,
                                        .000283*b, .0453*b, .00849*b, .225*b,
                                        .000102*b, .01632*b, .00306*b, .0812*b,
                                        .000614*b, .0984*b, .0184*b, .4887*b)
                              ) %>%
  mutate(count = round(count, 1))

indy_base_plot <- racial_baseline_indy %>%
  ggplot(aes(x = citizen_race, y = officer_race, fill = count, color = "black")) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "orange2") +
  geom_text(aes(label=count), size = 4, color = "black") + 
  theme(legend.position = "none") + 
  labs(x = "Citizen Race", y = "Officer Race", title = "Indianapolis Population Baseline")


# loss function

total <- 0

for (i in 1:16){
  o <- racial_data[i,3] %>%
    pull(1)
  e <- racial_baseline_uof[i,3] %>%
    pull(1)
  temp <- (o - e)
  total <- temp + total
}


expected_dst <- plot_grid(indy_base_plot, uof_base_plot, nrow = 1, ncol = 2)


clean_data %>%
  group_by(officer_number) %>%
  summarize(n = n()) %>%
  arrange(-n)


clean_dt <- clean_data %>%
  select(date, officer_number, district, street_num, street_name, type_of_force, citizen_arrested, citizen_hospitalized, officer_hospitalized, citizen_race, citizen_sex, officer_race, officer_sex)
```





